---
title: "Combine the microarray and RNA-seq data and generated masked data"
author: J. Taroni for ALSF CCDL
date: 10/12/2018
output: html_notebook
---

Because we need to mask values (e.g., they can't be missing tp start), we'll do 
an inner join for this particular experiment. 
This is different from how we expect to do imputation in production.

### Functions

```{r}
# magrittr pipe
`%>%` <- dplyr::`%>%`

# microarray data is "missing" a column name, we need to fill it in
read_in_microarray_data <- function(filename) {
  # read in a given file and set the first column name to "Gene"
  # we're suppressing warnings here because we know the missing column is
  # a problem -- that's why we have this function!
  df <- suppressWarnings(data.table::fread(filename, data.table = FALSE))
  colnames(df)[1] <- "Gene"
  return(df)
}
```

## Read in files and aggregate

First, the microarray data

```{r}
# full path to all the microarray data we obtain
microarray_files <- list.files(file.path("data", "microarray"), 
                               full.names = TRUE)
# initialize a list to hold the data.frames 
df_list <- list()
for (file_iter in seq_along(microarray_files)) {
  df_list[[file_iter]] <- read_in_microarray_data(microarray_files[file_iter])
}
```

Now the inner join.
We also have to get rid of the `_at` trailing the `ENSDARG` ids that is an
artifact of using Brainarray.

```{r}
dir.create(file.path("data", "aggregated"), recursive = TRUE)
microarray_df <- plyr::join_all(df_list, by = "Gene", type = "inner") %>%
  dplyr::mutate(Gene = sub("_at", "", Gene))
readr::write_tsv(microarray_df, file.path("data", "aggregated",
                                          "microarray_zebrafish.tsv"))
```

Now we'll repeat the process with RNA-seq data.

```{r}
seq_files <- list.files(file.path("data", "rnaseq"), full.names = TRUE)

df_list <- list()
for (file_iter in seq_along(seq_files)) {
  df_list[[file_iter]] <- data.table::fread(seq_files[file_iter],
                                            data.table = FALSE)
}

seq_df <- plyr::join_all(df_list, by = "Gene", type = "inner")
readr::write_tsv(seq_df, file.path("data", "aggregated", "seq_zebrafish.tsv"))
```

And join all the data together, again using an inner join because we'll need
to mask values.

```{r}
both_df <- dplyr::inner_join(microarray_df, seq_df, by = "Gene")
# are there any missing values
any(is.na(both_df))
```

Write the "complete" `data.frame` to file.

```{r}
readr::write_tsv(both_df, file.path("data", "aggregated", "all_zebrafish.tsv"))
```

Let's also quantile normalize the data and write that to file.
The rationale is that the _end result_ in a compendium will be quantile 
normalized and therefore we should use the quantile normalized expression 
values as our true values.

We need a matrix for quantile normalization.

```{r}
both_mat <- both_df %>%
  tibble::column_to_rownames("Gene") %>%
  as.matrix()
```

And we'll also need the target!

```{r}
target_df <- 
  readr::read_tsv(file.path("data", "qn_target", 
                            "d8gyazczr661umla1979zxu7_1538153041_target.tsv"),
                  col_names = FALSE)
target <- as.vector(as.matrix(target_df))
```

```{r}
# BiocInstaller::biocLite("preprocessCore")
qn_mat <- preprocessCore::normalize.quantiles.use.target(both_mat,
                                                         target = target)
qn_df <- data.frame(Gene = rownames(both_mat), qn_mat)
colnames(qn_df)[2:ncol(qn_df)] <- colnames(both_mat)
readr::write_tsv(qn_df, file.path("data", "aggregated", 
                                  "qn_all_zebrafish.tsv"))
```

We'll only keep around what we want for the masking bit.

```{r}
rm(list = setdiff(ls(), c("both_mat", "both_df", "%>%")))
```

## Masking

#### Functions

```{r}
mask_completely_at_random <- function(exprs_mat, indices_df,
                                      seed = 12345, perc_masked = 0.3) {
  # given an expression matrix (exprs_mat) and a "grid" of its indices 
  # (indices_df, where the row indices are in the first column and column 
  # indices are in the second column), mask some percentage (perc_masked)
  # of the values by replacing them with NAs
  
  # set seed for reproducibility!
  set.seed(seed)
  
  # how many cells do we want to mask in the expression data.frame and which
  # cells are they?
  ncells <- floor(nrow(indices_df) * perc_masked)
  rand_indices_df <- indices_df[sample(1:nrow(indices_df), ncells), ]
  
  # we need to return a matrix that has NAs -- the values are "masked"
  masked_exprs_mat <- exprs_mat
  masked_exprs_mat[as.matrix(rand_indices_df)] <- NA
  
  # We also want to store a data.frame that includes the row and column indices
  # of the values that were masked and the original values. This facilitates
  # our downstream error calculations. 
  rand_indices_df$original_values <- exprs_mat[as.matrix(rand_indices_df)]

  # return the masked expression matrix (will passed to imputer of
  # choice), the data.frame that keeps track of indices and original values,
  # and the seed used to generate them
  return_list <- list(exprs_mat = masked_exprs_mat,
                      indices = rand_indices_df,
                      seed = seed)
  return(return_list)
  
}

mask_random_rows <- function(exprs_mat, seed = 12345) {
  # 10% of rows missing 10% of values
  # 10% of rows missing 20% of values
  # 5% of rows missing 30% of values
  
  get_masked_columns <- function(expression_mat, perc_masked = 0.1) {
    # 10% of values will be masked by default
    col_indices <- sample(1:ncol(expression_mat), 
                          floor(ncol(expression_mat) * perc_masked))
  }
  
  get_indices_df <- function(expression_matrix, 
                             row_indices, 
                             percent_masked) {
    # for each row index, get an appropriate level of masked columns/samples
    column_index_list <- lapply(row_indices, 
                                function(x) 
                                  get_masked_columns(expression_matrix,
                                                     percent_masked))
    names(column_index_list) <- as.character(row_indices)
    # we want this in data.frame form
    indices_df <- reshape2::melt(column_index_list)
    # reshape makes it that the elements of the list are in the first column
    # these are the column indices in this case
    colnames(indices_df) <- c("column", "row")
    # and that's the opposite of the order we want
    indices_df <- indices_df[, c("row", "column")]
    return(indices_df)
  }
  
  # set seed for reproducibility
  set.seed(seed)
  
  # all rows are able to be masked at this point, but every time a row gets 
  # selected to be masked in a "round" with fewer missing values, take it
  # out of the running
  available_rows <- 1:nrow(exprs_mat)
  # the 10% of rows that will be missing 10% of values
  ten_perc_rows <- sample(available_rows, floor(nrow(exprs_mat) * 0.1))
  # take these out of the pool of rows to mask
  available_rows <- available_rows[-which(available_rows %in% ten_perc_rows)]
  # repeat for 10% of rows missing 20% of values
  twenty_perc_rows <- sample(available_rows, floor(nrow(exprs_mat) * 0.1))
  available_rows <- available_rows[-which(available_rows %in% twenty_perc_rows)]
  # only 5% will be missing 30% of values
  thirty_perc_rows <- sample(available_rows, floor(nrow(exprs_mat) * 0.05))
  
  # let's get an indices data.frame for each of these guys -- we'll bind the
  # rows together and then basically repeat the MCAR procedure above
  ten_indices_df <- get_indices_df(expression_matrix = exprs_mat,
                                   row_indices = ten_perc_rows,
                                   percent_masked = 0.1)
  twenty_indices_df <- get_indices_df(expression_matrix = exprs_mat,
                                      row_indices = twenty_perc_rows,
                                      percent_masked = 0.2)
  thirty_indices_df <- get_indices_df(expression_matrix = exprs_mat,
                                      row_indices = thirty_perc_rows,
                                      percent_masked = 0.3)
  
  # one indices data.frame to rule them all and make sure the values are 
  # integers
  indices_df <- rbind(ten_indices_df, twenty_indices_df, thirty_indices_df)
  indices_df$row <- as.integer(indices_df$row)
  indices_df$column <- as.integer(indices_df$column)
  
  # we need to return a matrix that has NAs -- the values are "masked"
  masked_exprs_mat <- exprs_mat
  masked_exprs_mat[as.matrix(indices_df)] <- NA
  
  # We also want to store a data.frame that includes the row and column indices
  # of the values that were masked and the original values. This facilitates
  # our downstream error calculations. 
  indices_df$original_values <- exprs_mat[as.matrix(indices_df)]
  # let's also keep track of how many values were missing in this row, I think
  # this will help with plotting later
  indices_df$missingness <- c(rep(0.1, nrow(ten_indices_df)),
                              rep(0.2, nrow(twenty_indices_df)),
                              rep(0.3, nrow(thirty_indices_df)))
  
  # return the masked expression matrix (will passed to imputer of
  # choice), the data.frame that keeps track of indices and original values,
  # and the seed used to generate them
  return_list <- list(exprs_mat = masked_exprs_mat,
                      indices = indices_df,
                      seed = seed)
  return(return_list)
  
}
```

We're going to repeat these masking procedures ten times, each with different
random seeds.

```{r}
set.seed(12345)
seeds <- sample(1:10000, 10)
```

First, the missing completely at random examples

```{r}
# we need all the indices in the expression matrix 
mcar_indices_df <- expand.grid(1:nrow(both_mat), 1:ncol(both_mat))

# 30% values missing completely at random
mcar_list <- 
  lapply(seeds,
         function(seed) mask_completely_at_random(exprs_mat = both_mat,
                                                  indices_df = mcar_indices_df,
                                                  seed = seed, 
                                                  perc_masked = 0.3))
names(mcar_list) <- as.character(seeds)
```

And now the "more realistic example"

```{r}
random_rows_list <- 
  lapply(seeds,
         function(seed) mask_random_rows(exprs_mat = both_mat,
                                         seed = seed))
names(random_rows_list) <- as.character(seeds)
```

Let's save these lists as `.RDS`.
We'll prep this data for use with [`KNNImputer`](https://libsleipnir.bitbucket.io/KNNImputer.html)
in another script that includes different processing.

```{r}
masked_dir <- file.path("data", "masked") 
dir.create(masked_dir, recursive = TRUE, showWarnings = FALSE)
saveRDS(mcar_list, file = file.path(masked_dir, "mcar_list.RDS"))
saveRDS(random_rows_list, file = file.path(masked_dir, 
                                           "missing_random_rows_list.RDS"))
```

