---
title: "Processing data for imputation four ways"
author: J. Taroni for ALSF CCDL
date: 10/17/2018
output: html_notebook
---

We intend to combine microarray and RNA-seq data in our species compendia and
we need to select an (initial) processing pipeline for our imputation.

So, here's what we're going to test:

* Splitting by technology (e.g., imputing microarray and RNA-seq data 
separately), performing no transformations
* Splitting by technology, `log2(x + 1)` transformation of RNA-seq data
* Imputing technologies together, no transformation
* Imputing technologies together, `log2(x + 1)` transformation of RNA-seq data
* Imputing technologies, all samples are quantile normalized

The rationale behind the log2 transformation is to reduce the effect of outliers
on the KNN imputation step which will use Euclidean distance.
By imputing technologies together, you have more samples, which has been shown
to be helpful in prior work and is likely to be helpful here.

In `03-aggregate_and_mask`, we generated missing completely at random data
and a somewhat more realistic structure (we think) where 25% of rows have a
"significant" portion of values missing (10%, 20%, or 30%) for 10 repeats
with different random seeds.

```{r}
write_five_pcl <- function(exprs_mat, target, path, seed, 
                           missing_type = "MCAR") {
  # given an expression matrix, let's get 5 PCL files that satisfy the above
  # conditions; file names TBD by the random seed used and the missing type
  # 
  #   all_un: both technologies, no transformation
  #   all_log: both technologies, RNA-seq is log2(x + 1)
  #   microarray: microarray data only, no transformation
  #   seq_log: RNA-seq data only, log2(x + 1)
  #   seq_un: RNA-seq data only, no transformation

  # the Ensembl gene IDs are the rownames, we'll want them in their own 
  # column
  exprs_df <- tibble::rownames_to_column(data.frame(exprs_mat), var = "Gene")
  # write both technologies, no transformation to file
  readr::write_tsv(exprs_df, file.path(path, paste0("all_un_", seed, "_", 
                                                    missing_type, ".pcl")))
  
  # the sample names tell us where the sample came from and therefore the
  # technology 
  seq_sample_patterns <- "DRR|ERR|SRR"
  
  # split by technology
  microarray_df <- exprs_df[, -grep(seq_sample_patterns, colnames(exprs_df))]
  seq_df <- exprs_df[, c(1, grep(seq_sample_patterns, colnames(exprs_df)))]
  
  # write the individual technologies to file
  readr::write_tsv(microarray_df, file.path(path, 
                                            paste0("microarray_", seed, "_", 
                                                   missing_type, ".pcl")))
  readr::write_tsv(seq_df, file.path(path, paste0("seq_un_", seed, "_", 
                                                  missing_type, ".pcl")))  
  
  # now let's log transform the seq data only
  log_seq_df <- data.frame(Gene = seq_df$Gene,
                           log2(seq_df[, 2:ncol(seq_df)] + 1))
  readr::write_tsv(log_seq_df, file.path(path, paste0("seq_log_", seed, "_", 
                                                      missing_type, ".pcl")))
  
  # both technologies, with the RNA-seq data log transformed
  all_log_df <- suppressMessages(dplyr::inner_join(microarray_df, log_seq_df, 
                                                   by = "Gene"))
  readr::write_tsv(all_log_df, file.path(path, paste0("all_log_", seed, "_", 
                                                      missing_type, ".pcl")))
  
  # both technologies, quantile normalized
  all_log_mat <- as.matrix(tibble::column_to_rownames(all_log_df, "Gene"))
  all_log_qn_mat <- preprocessCore::normalize.quantiles.use.target(all_log_mat,
                                                                   target)
  all_qn_df <- data.frame(Gene = all_log_df$Gene, all_log_qn_mat)
  colnames(all_qn_df) <- colnames(all_log_df)
  readr::write_tsv(all_qn_df, file.path(path, paste0("all_qn_", seed, "_", 
                                                      missing_type, ".pcl")))
  
  # don't return all_log_df
  return(NULL)
}
```

### Target for quantile normalization

```{r}
target_df <- 
  readr::read_tsv(file.path("data", "qn_target", 
                            "d8gyazczr661umla1979zxu7_1538153041_target.tsv"),
                  col_names = FALSE)
target <- as.vector(as.matrix(target_df))
rm(target_df)
```

### Missing completely at random

```{r}
mcar_dir <- file.path("data", "masked", "MCAR")
dir.create(mcar_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r}
mcar_list <- readRDS(file.path("data", "masked", "mcar_list.RDS"))
```

```{r}
mcar_pcl <- lapply(mcar_list, 
                   function(x) write_five_pcl(exprs_mat = x$exprs_mat,
                                              target = target,
                                              path = mcar_dir,
                                              seed = x$seed,
                                              missing_type = "MCAR"))
rm(mcar_pcl, mcar_dir, mcar_list)
```

### Missing random rows

```{r}
mrr_dir <- file.path("data", "masked", "MRR")
dir.create(mrr_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r}
mrr_list <- readRDS(file.path("data", "masked", 
                              "missing_random_rows_list.RDS"))
mrr_pcl <- lapply(mrr_list, 
                   function(x) write_five_pcl(exprs_mat = x$exprs_mat,
                                              target = target,
                                              path = mrr_dir,
                                              seed = x$seed,
                                              missing_type = "MRR"))
```

