---
title: "PCA on zebrafish test compendium"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: J. Taroni for CCDL
date: 2019
---

Here, we're using a large zebrafish test compendium that was run in production.
It's majority RNA-seq data.
We had early results using a small test compendium that suggested there was a 
notable "platform bias" when using PCA.
We'll perform PCA on the large compendium in this notebook with particular
regard for technology.

## Setup

#### Packages and functions

```{r}
library(ggfortify)
```

```{r}
`%>%` <- dplyr::`%>%`
```

#### Directories/files

```{r}
test_compendium_dir <- file.path("refinebio-data", "larger-compendium", 
                                 "DANIO_RERIO")
compendium_exprs_file <- file.path(test_compendium_dir, "DANIO_RERIO.tsv")
```

Directory for plots

```{r}
plot_dir <- "plots"
dir.create(plot_dir, showWarnings = FALSE)
```

## Main

Reading in the compendium expression data

```{r}
compendium_exprs_df <- readr::read_tsv(compendium_exprs_file,
                                       progress = FALSE)
```

We can _infer_ which technology a sample is "part of" from the column names
of this `data.frame`.

```{r}
set.seed(123)
colnames(compendium_exprs_df)[sample(2:ncol(compendium_exprs_df), 6)]
```

RNA-seq samples will have `ERR`, `SRR`, or `DRR` in their identifiers (column
names).
Let's generate _labels_ for the samples using this information.

```{r}
sample_names <- colnames(compendium_exprs_df)[-1]  # don't want gene id column
compendium_technology_labels <- rep("MICROARRAY", length(sample_names))
compendium_technology_labels[grep("ERR|DRR|SRR", sample_names)] <- "RNA-SEQ"
```

Get expression values as a matrix 

```{r}
exprs_mat <- compendium_exprs_df %>%
  as.data.frame() %>%
  tibble::column_to_rownames("X1") %>%
  as.matrix()
saveRDS(exprs_mat, file.path("data", "larger_compendium_exprs_mat.RDS"))
```

### PCA

```{r}
pca_results <- prcomp(t(exprs_mat))
```

Prepare as `data.frame` for use elsewhere

```{r}
pc_df <- pca_results$x[, 1:2] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Sample")
```

```{r}
# get the technology labels as a data.frame for plotting
technology_df <- as.data.frame(compendium_technology_labels) 
```

```{r}
pc_df <- technology_df %>%
  dplyr::mutate("Sample" = sample_names) %>%
  dplyr::inner_join(y = pc_df, by = "Sample")

readr::write_tsv(pc_df, path = file.path("results",
                                         "larger_compendium_PC1and2.tsv"))
```

### Plotting

```{r}
# plotting with autoplot for PCA
autoplot(pca_results, data = technology_df, alpha = 0.5, 
         colour = 'compendium_technology_labels', scale = 0) +
  scale_color_manual(labels = c("MICROARRAY", "RNA-SEQ"), 
                     values = c("#E69F00", "#56B4E9")) +
  theme_bw() +
  ggtitle("Larger Test Compendium by Technology") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  guides(colour = guide_legend(title = "technology"))
```

I'll note here that because of the way we process RNA-seq data in refine.bio,
these RNA-seq samples with _necessarily_ be part of the same experiments. 
There are no such guarantees for the microarray samples.
We can see that there are some "clumps" in the RNA-seq, which, if I had to 
guess, likely correspond to samples from an individual experiment.
_Here, I'm using the term "experiment" to refer to what is called a study (e.g., 
`SRP`, `ERP`, or `DRP`) in SRA._

Save the plot to file.

```{r}
ggsave(filename = file.path(plot_dir, "larger_test_compendium_PCA.png"),
       plot = last_plot(), width = 7, height = 5)
```

```{r}
png(file.path(plot_dir, "larger_test_compendium_PCA_pairs_plot.png"))
pairs(pca_results$x[, 1:5], col = dplyr::recode(compendium_technology_labels,
                                                `MICROARRAY` = "#E69F00",
                                                `RNA-SEQ` = "#56B4E9"))
dev.off()
```

## Session info

```{r}
sessionInfo()
```
