---
title: "Zebrafish gene lengths"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: J. Taroni for CCDL
date: 2019
---

In `7-technology_diff_exp`, we observed that genes that are shorter tend to have
lower expression values in RNA-seq data in the test compendium.

Here, we're interested in if short genes tend to be unobserved (e.g., have
counts/count-scale data that are zero) in the RNA-seq data used in the test 
compendium.
We'll go back to the `lengthScaledTPM.tsv` that has not yet been through
the aggregation process to explore this.

## Set up

```{r}
`%>%` <- dplyr::`%>%`
```

```{r}
# we'll need ggfortify to do the PCA plots
library(ggfortify)
```

```{r}
# directory to hold the plots
plot_dir <- "plots"
```

#### Compendium genes

We will explore this gene lengths-observations/expression levels relationship in
_all_ genes and also subset to only the genes in the test compendium, so we 
need to pull those gene identifiers out of the test compendium expression
matrix.

```{r}
compendium_exprs_file <- file.path("refinebio-data", "DANIO_RERIO", 
                                   "DANIO_RERIO.tsv")
compendium_genes <- readr::read_tsv(compendium_exprs_file, progress = FALSE) %>%
  dplyr::pull(X1)
```

#### Read in `lengthScaledTPM.tsv` files

```{r}
list_of_rnaseq_files <- list.files(file.path("..", "data", "rnaseq/"),
                                   full.names = TRUE)
```

Read in the files such that they are individual `data.frame` in a list

```{r}
rnaseq_df_list <- lapply(list_of_rnaseq_files, readr::read_tsv)
```

```{r}
# bind all the individual sample data.frame together
rnaseq_df <- dplyr::bind_cols(rnaseq_df_list)
# we end up with a bunch of "extra" gene columns this way, but they should be
# in the same order
all.equal(rnaseq_df$Gene1, rnaseq_df$Gene112)
```

```{r}
# drop the extra gene columns and set the rownames to gene identifiers -- this
# will help with calculating the proportion of zeros, etc.
rnaseq_df <- 
  rnaseq_df[, c(TRUE, !grepl("Gene", colnames(rnaseq_df))[2:ncol(rnaseq_df)])]
rnaseq_df <- rnaseq_df %>%
  as.data.frame() %>%
  tibble::column_to_rownames("Gene")
# remove the list of individual sample data.frame
rm(rnaseq_df_list)
```

```{r}
# read in gene lengths
gene_lengths_df <- 
  readr::read_tsv(file.path("data", 
                            "Danio_rerio.GRCz11_gene_lengths_exons.tsv"))
```

## Gene zero counts and expression levels

### For each gene, how many samples have a count-scale expression value of zero?

```{r}
# calculate the proportion of samples a gene is zero for
gene_zeroes_proportion <- data.frame(
  "Gene" = rownames(rnaseq_df),
  "Zero_proportion" = apply(rnaseq_df, 1, 
                            function(x) sum(x == 0) / length(x))
)
```

```{r}
# bind to lengths, this will be used for plotting the relationship between
# gene length and number of zeroes
zeroes_df <- dplyr::inner_join(gene_lengths_df, gene_zeroes_proportion,
                               by = "Gene")
```

```{r}
# there are some really long genes that will probably make it difficult to
# visualize what is going on
summary(zeroes_df$gene_length)
```

#### All genes

```{r}
# filtering genes based on length to get a clearer picture
zeroes_df %>%
  dplyr::filter(gene_length <= 20000) %>%
  ggplot(aes(x = gene_length, y = Zero_proportion)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  theme_bw() +
  labs(x = "gene length", y = "proportion of samples with zero count",
       subtitle = "All genes", title = "Gene length and observations")
```

```{r}
ggsave(filename = file.path(plot_dir, 
                            "gene_length_vs_zero_count_all_genes.png"),
       plot = last_plot())
```

#### Compendium genes only

```{r}
# subset just to genes that are in the compendium
zeroes_df %>%
  dplyr::filter(Gene %in% compendium_genes, 
                gene_length <= 20000) %>%
  ggplot(aes(x = gene_length, y = Zero_proportion)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  theme_bw() +
  labs(x = "gene length", y = "proportion of samples with zero count",
       subtitle = "Test compendium genes only", 
       title = "Gene length and observations")
```

```{r}
ggsave(filename = file.path(plot_dir, 
                            "gene_length_vs_zero_count_compendium_genes.png"),
       plot = last_plot())
```

#### Takeaway

Shorter genes do tend to have a larger proportion of samples where they have
zero values. 
It stands to reason that if we do not _observe_ shorter genes, we can not
correct for length-bias effective. 
(Recall that `lengthScaledTPM` are calculated using the average tx length
observed across samples.)

### Removing "short" genes from the test species compendium

Does removing shorter genes from the species compendium remove evidence of a
technology bias?

```{r}
# read in the compressed version of the test species compendium
exprs_mat_file <- file.path("data", "refinebio_test_compendium_exprs_mat.RDS")
exprs_mat <- readRDS(exprs_mat_file)
```

```{r}
# derive the technology labels from the sample names
sample_names <- colnames(exprs_mat)
compendium_technology_labels <- rep("MICROARRAY", length(sample_names))
compendium_technology_labels[grep("lengthScaledTPM", sample_names)] <- "RNASEQ"
```

```{r}
# what genes are short? using 
short_genes <- zeroes_df %>%
  dplyr::filter(gene_length <= quantile(zeroes_df$gene_length, 0.25)) %>%
  dplyr::pull(Gene)
```
Drop the "short" genes from the expression matrix

```{r}
length_filtered_exprs_mat <- exprs_mat[!rownames(exprs_mat) %in% short_genes, ]
dim(length_filtered_exprs_mat)
```

#### PCA

PC1 and PC2

```{r}
# PCA on the length filtered exprs mat
pca_results <- prcomp(t(length_filtered_exprs_mat))

# get the technology labels as a data.frame for plotting
technology_df <- as.data.frame(compendium_technology_labels) 

# plotting with autoplot for PCA
autoplot(pca_results, data = technology_df, alpha = 0.5, 
         colour = 'compendium_technology_labels', scale = 0) +
  scale_color_manual(labels = c("MICROARRAY", "RNA-SEQ"), 
                     values = c("#E69F00", "#56B4E9")) +
  theme_bw() +
  ggtitle("Test Compendium by Technology: Short genes removed") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  guides(colour = guide_legend(title = "technology"))
```

```{r}
ggsave(filename = file.path(plot_dir, 
                            "gene_lengths_test_compendium_PCA.png"),
       plot = last_plot())
```

Pairs plot (PC1-5)

```{r}
png(file.path(plot_dir, "gene_lengths_test_compendium_pairs_plot.png"))
pairs(pca_results$x[, 1:5], col = dplyr::recode(compendium_technology_labels,
                                                `MICROARRAY` = "#E69F00",
                                                `RNASEQ` = "#56B4E9"),
      main = "Test Compendium by Technology: Short genes removed")
dev.off()
```

#### Takeaway

Removing short genes from the test species compendium does not remove 
technology bias, but there are some considerations:

* These data have already been quantile normalized -- if the gene length
contributes to the expression values prior to quantile normalization, this 
bias could possibly "leak into" the quantile normalized values for all the
genes.

## Session Info

```{r}
sessionInfo()
```

