---
title: "Construct _Danio rerio_ prior information matrices"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: J. Taroni for CCDL
date: 2019
---

If we want to use `PLIER`, we need to construct prior information matrices
for zebrafish.
We'll do that using two sources: 
[WikiPathways](https://www.wikipathways.org/index.php/WikiPathways) and 
[ZFIN](https://zfin.org/)

## Setup

```{r}
`%>%` <- dplyr::`%>%`
```

```{r}
# BiocManager::install("rWikiPathways", "GSA", "org.Dr.eg.db", update = FALSE)
# install.packages("XML")
library(rWikiPathways)
library(XML)
library(org.Dr.eg.db)
```

```{r}
pathway_dir <- file.path("data", "pathways")
dir.create(pathway_dir, showWarnings = FALSE, recursive = TRUE)
```

## WikiPathways

```{r}
# download current version of WikiPathway pathways
# this is CC0
downloadPathwayArchive(organism = "Danio rerio", format = "gmt",
                       destpath = pathway_dir, date = "20190110")
```

Read the `.gmt` file in and convert to binary matrix

```{r}
zebrafish_wp_list <- 
  GSA::GSA.read.gmt(file.path(pathway_dir, 
                              "wikipathways-20190110-gmt-Danio_rerio.gmt"))
```

Clean up the list to make it more amenable to the binary prior information
matrix we'll need for use with PLIER.

```{r}
names(zebrafish_wp_list$genesets) <- make.names(  # remove hyphens
  gsub(" ", "_",  # convert spaces to underscores
       gsub("%WikiPathways_20190110%", " ",  # remove date, WP information 
            gsub("%Danio rerio", "",  # remove species information
                 zebrafish_wp_list$geneset.names))
  )
)
# remove pathway names and descriptions
zebrafish_wp_list <- zebrafish_wp_list$genesets
```

We're going to use ZFIN identifiers instead of Entrez IDs, which are what are
used in WikiPathways.

```{r}
# our goal is to get a binary matrix that represents gene-pathway relationships
# to do this, we first need to melt the list into a data.frame
zebrafish_wp_binary_df <- reshape2::melt(zebrafish_wp_list) %>%
  # get sensible column names
  dplyr::mutate(Gene = value, Pathway = L1) %>%
  dplyr::select(Gene:Pathway) %>%
  # conversion from Entrez IDs to ZFIN IDs with mapIds -- this should be 1:1
  dplyr::mutate(ZFIN = mapIds(org.Dr.eg.db, 
                              keys = as.character(Gene), 
                              keytype = "ENTREZID", column = "ZFIN")) %>%
  # drop Entrez IDs
  dplyr::select(-Gene) %>%
  # drop records missing ZFIN ids
  dplyr::filter(!is.na(ZFIN)) %>%
  # unique records only, just in case
  dplyr::distinct() %>%
  # get binary matrix, where the first column will be gene and the columns are
  # pathways
  reshape2::dcast(ZFIN ~ Pathway, fun.aggregate = length)
```

```{r}
# now a more PLIER friendly format
zebrafish_wp_binary_mat <- zebrafish_wp_binary_df %>%
  tibble::column_to_rownames("ZFIN") %>%
  as.matrix()
saveRDS(zebrafish_wp_binary_mat, 
        file.path(pathway_dir, "Danio_rerio_wikipathways_prior_info.RDS"))
```

## ZFIN

Given the [licensing of the ZFIN resource](https://zfin.org/warranty.html) and 
the fact that downloading via the web interface has the advantage of including 
the archive date in the file name/header and including the column names, we've 
obtained these files via the 
[downloads webpage](https://zfin.org/downloads/archive/2019.01.25) using the
date that is currently the most recent archive (`2019-01-25`).


```{r}
wildtype_expression_file <- 
  file.path(pathway_dir, "wildtype-expression_fish_2019.01.25.txt")
# we're only going to use the in situ data
wildtype_expression_df <- readr::read_tsv(wildtype_expression_file, skip = 1) %>%
  dplyr::filter(Assay == "mRNA in situ hybridization") %>%
  dplyr::select(-X15)
```

We're hoping to get two gene sets from this ZFIN information: 1) anatomical
structure and 2) developmental stage.
Parsing this by stage-tissue pairs ends up being extremely sparse, so we can't
approach it that way and get anything useful.

### Stages

```{r}
stage_file <- file.path(pathway_dir, "stage_ontology_2019.01.25.txt")
stage_df <- readr::read_tsv(stage_file, skip = 1) %>%
  # drop column associated with parsing error
  dplyr::select(-X6)
```

```{r}
stage_ordered_factor <- factor(stage_df$`Stage Name`[-1],  # drop unknown
                               levels = stage_df$`Stage Name`[-1],
                               ordered = TRUE)
stage_ordered_factor
```
```{r}
gene_stages_list <- list()
for (current_gene in unique(wildtype_expression_df$`Gene ID`)) {
  
  # initialize vector to hold _all_ the stages for the current gene
  current_gene_stages <- vector()
  
  # data.frame filtered to current gene only
  current_gene_df <- wildtype_expression_df %>%
    dplyr::filter(`Gene ID` == current_gene)
  
  # all rows in current_gene_df
  for (gene_row_iter in 1:nrow(current_gene_df)) {
    # pull pertinent stage information from this row
    start_stage <- current_gene_df$`Start Stage`[gene_row_iter]
    end_stage <- current_gene_df$`End Stage`[gene_row_iter]
    
    # find the indices corresponding to the stages and extract the stage
    # names from the ordered factor vector
    start_index <- which(stage_ordered_factor == start_stage)
    end_index <- which(stage_ordered_factor == end_stage)
    covered_stages <- as.character(stage_ordered_factor[start_index:end_index])
    current_gene_stages <- append(current_gene_stages,
                                  covered_stages)
  }
  gene_stages_list[[current_gene]] <- current_gene_stages
}
```

We want to go from this list to a binary matrix for use with PLIER

```{r}
gene_stages_binary_mat <- reshape2::melt(gene_stages_list) %>%
  # remove any duplicate stages
  dplyr::distinct() %>%
  # sensible column names
  dplyr::mutate(Stage = value, Gene = L1) %>%
  dplyr::select(Stage:Gene) %>%
  # and finally, get a PLIER-friendly binary matrix
  reshape2::dcast(Gene ~ Stage, fun.aggregate = length) %>%
  tibble::column_to_rownames("Gene") %>%
  as.matrix()
# save to file
saveRDS(gene_stages_binary_mat,
        file = file.path(pathway_dir, 
                         "ZFIN_stages_prior_info.RDS"))
```

### Tissues

We'll use the adult stage _only_ to do the gene-tissue mapping.

```{r}
# filter to only adult
adult_wt_df <- wildtype_expression_df %>%
  dplyr::filter(`End Stage` == "Adult")

# then we can get a list by tissue -> binary matrix
adult_tissue_wt_binary_mat <- 
  reshape2::melt(
    # this gives us a list that we're melting into a data.frame
    split(adult_wt_df$`Gene ID`,
          adult_wt_df$`Super Structure Name`)) %>%
  # setting informative column names
  dplyr::mutate(Gene = value, Tissue = L1) %>%
  dplyr::select(Gene:Tissue) %>%
  # filtering out any duplicate records
  dplyr::distinct() %>%
  # and finally, get a PLIER-friendly binary matrix
  reshape2::dcast(Gene ~ Tissue, fun.aggregate = length) %>%
  tibble::column_to_rownames("Gene") %>%
  as.matrix()

# save the RDS to file
saveRDS(adult_tissue_wt_binary_mat, 
        file.path(pathway_dir, "ZFIN_adult_tissue_prior_info.RDS"))
```

## Session info

```{r}
sessionInfo()
```

