---
title: "Run PLIER on test compendium"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: J. Taroni for CCDL
date: 2019
---

Do we "get more biology" out of the combined technology test compendium than we
get out of each of the technologies individually?

We'll use the metrics we developed as part of the MultiPLIER project here.
Specifically, do we get higher pathway coverage for a model trained on
the dual-technology test compendium vs. those trained on the individual
technology compendia?

## Setup

#### Functions

Source the most recent version of the MultiPLIER utils from Github

```{r}
source("https://github.com/greenelab/multi-plier/raw/7c56a2867f9f08f7bcd3617875c7c2dbe886cbeb/util/plier_util.R")
# we only want GetPathwayCoverage
rm(list = setdiff(ls(), "GetPathwayCoverage"))
```

```{r}
`%>%` <- dplyr::`%>%`
```

```{r}
# we'll train 3 PLIER models: matrix with both technologies, microarray samples
# only, and RNA-seq samples only
PLIERWrapper <- function(exprs_mat, common_genes, pathway_mat, seed = 123) {
  
  # set seed for reproducibility
  set.seed(seed)
  
  # z-score the expression data
  row_norm_exprs <- PLIER::rowNorm(exprs_mat[common_genes, ])
  
  # how many "significant PCs" via the elbow method -- we use this to set k when
  # training a PLIER model
  num_pcs <- PLIER::num.pc(row_norm_exprs)
  
  return(PLIER::PLIER(data = row_norm_exprs[common_genes, ], 
                      priorMat = pathway_mat[common_genes, ],
                      k = round(num_pcs * 1.3)))
}
```

#### Libraries

```{r}
# PLIER installation
# BiocManager::install("qvalue", update = FALSE)
# devtools::install_github('wgmao/PLIER', 
#                          ref = 'a2d4a2aa343f9ed4b9b945c04326bebd31533d4d', 
#                          dependencies = TRUE)
```

```{r}
library(PLIER)
library(org.Dr.eg.db)
```

#### Directories

```{r}
pathway_dir <- file.path("data", "pathways")
result_dir <- "results"
plot_dir <- "plots"
```

Results directory specifically for holding the PLIER models

```{r}
plier_result_dir <- file.path(result_dir, "PLIER_models")
dir.create(plier_result_dir, showWarnings = FALSE, recursive = TRUE)
```

## Read in data

### Prior information matrices

We need to get the 3 pathway prior information matrices we have: developmental
stage, tissues (based on adult only), and the wikipathways into a single 
matrix.
We can do this with the `PLIER::combinePaths` function.

```{r}
# read in the three individual matrices we prepped in 
# 09-prior_information_matrix
wikipathway_mat <- readRDS(file.path(pathway_dir, 
                                     "Danio_rerio_wikipathways_prior_info.RDS"))
adult_tissue_mat <- readRDS(file.path(pathway_dir,
                                      "ZFIN_adult_tissue_prior_info.RDS"))
dev_stage_mat <- readRDS(file.path(pathway_dir,
                                   "ZFIN_stages_prior_info.RDS"))
# combine into a single matrix for use with PLIER
combine_path_mat <- PLIER::combinePaths(adult_tissue_mat,
                                        dev_stage_mat,
                                        wikipathway_mat)
# we no longer need the individual matrices
rm(adult_tissue_mat, dev_stage_mat, wikipathway_mat)
```

### Test compendium

We need to convert the test compendium gene identifiers from Ensembl gene IDs to
ZFIN identifiers so that the IDs in the prior information and the expression
data match.

```{r}
# first we need to read in the 
exprs_file <- file.path("data", "refinebio_test_compendium_exprs_mat.RDS")
exprs_mat <- readRDS(exprs_file)
```

We expect that there will be a 1:many mapping here, so we'll use the option
where only the first thing that comes back is returned by setting 
`multiVals = "first"` (note that this is the default behavior).

```{r}
# ENSDARG -> ZDB-GENE
zfin_ids <- mapIds(org.Dr.eg.db, keys = rownames(exprs_mat),
                   column = "ZFIN", keytype = "ENSEMBL",
                   multiVals = "first")

# these should be in the same order, but out of an abundance of caution!
if (all.equal(rownames(exprs_mat), names(zfin_ids))) {
  rownames(exprs_mat) <- zfin_ids
} else {
  stop("Something went wrong!")
}
```

## PLIER

### Both technologies

```{r}
# find the common genes between the expression matrix and the prior information
# matrix
common_genes <- PLIER::commonRows(exprs_mat, combine_path_mat)

# this expression matrix includes samples from both technologies
both_tech_plier <- PLIERWrapper(exprs_mat = exprs_mat,
                                common_genes = common_genes,
                                pathway_mat = combine_path_mat)
```

Check the pathway coverage for this model

```{r}
both_tech_pathway_coverage <- GetPathwayCoverage(both_tech_plier)
```

### Microarray only

```{r}
microarray_mat <- exprs_mat[, grep(".PCL", colnames(exprs_mat))]
microarray_plier <- PLIERWrapper(exprs_mat = microarray_mat,
                                 common_genes = common_genes,
                                 pathway_mat = combine_path_mat)
microarray_pathway_coverage <- GetPathwayCoverage(microarray_plier)
```

### RNA-seq only

```{r}
rnaseq_mat <- exprs_mat[, grep("lengthScaledTPM", colnames(exprs_mat))]
rnaseq_plier <- PLIERWrapper(exprs_mat = rnaseq_mat,
                             common_genes = common_genes,
                             pathway_mat = combine_path_mat)
rnaseq_pathway_coverage <- GetPathwayCoverage(rnaseq_plier)
```

#### Save models to file

```{r}
saveRDS(both_tech_plier, file = file.path(plier_result_dir,
                                          "both_technologies_PLIER.RDS"))
saveRDS(microarray_plier, file = file.path(plier_result_dir,
                                           "microarray_PLIER.RDS"))
saveRDS(rnaseq_plier, file = file.path(plier_result_dir,
                                       "rnaseq_PLIER.RDS"))
```

## Session info

```{r}
sessionInfo()
```

